Experience in Modeling a Microcontroller Instruction Set Using B

### 2012-05-20 trunk/papers/ArticleZ80/source/paper.tex
Versao submetida.

\title{Experience in Modeling a Microcontroller Instruction Set Using B}
%
\hyphenation{se-pa-ra-te}
\hyphenation{in-struc-tion}
\hyphenation{re-gis-ter}
\hyphenation{re-gis-ters}

@@

  This paper presents a formal model, developed in B, of the Z80
  instruction set. The formal model includes different libraries that
  specify entities commonly used in microcontrolers and
  microprocessors. Not only this model can be used for documentation
  purposes, but also to simulate and test the assembly code. It is also
  a fundamental component to extend the reach of the B method from the
  specification of the functional requirements of a system to an assembly-level
  implementation thereof.
% Using this model-driven approach and the instruction set model, a small case
% study to analyse a petroleum production test system in oil field was developed.
%  \paragraph[Keywords]{ \textbf{Keywords}: Embedded Software, Simulation, B
% method and Verification%}
\end{abstract}

@@

The new model supports animation in ProB~\cite{proB}.  Essentially, the
new model\footnote{The interested reader in more details is invited to visit our repository at: http://code.google.com/p/b2asm/.}
no longer contains definitions derived from infinite sets. Using B
constructs for modularity, auxiliary libraries of basic modules were
developed as part of the construction of the microcontroller model. Such
a library has many definitions of concepts common to microcontrollers.
Besides the new Z80 model, a petroleum production test system was
developed using the Z80 assembly language to analyse the code simulation
and the verification process.

@@

%TODO:[NEW PARAGRAPH-TO-REVIEW]
Similarly, several other general functions and lemmas were created for all other data types.
Some times, we need  create arithmetic and logic functions that are more specific for a
determined microprocessor, like Z80 that is presented in the section \ref{sec:z80}.

%TODO:[NEW PARAGRAPH-TO-REVIEW]
The definitions, functions and properties from hardware library and types library are difficult to verify,
on the other hand  these libraries can be reused in several others  microprocessors without to verify again.
Nearly 50\% of all these proof obligations shown in the Table~\ref{tab:types} are verified automatically 
or with few user pass\footnote{The user pass are steps that aid the prover to find the proof.} .
But many others proof obligations are difficult to verify, because they involve functions
that contain arithmetic expressions that are complex to manipulate with the interactive
prover of AtelierB.  However, all theses 531 proof obligations were verified completely.

@@

\begin{figure}[he]
\centering
\includegraphics[width=3.in]{images/Diagram_Types_and_Rules.pdf}
\caption{Conversion functions}
\label{DiagramTypesRules}
\end{figure}

@@

There is a specific function to refresh the flag register, it is named $\textit{update\_reg\_flag}$. It is typed and defined as follow:\it update\_flag\_reg \rm $\in$ \rm (\it BIT  $\times$  \it BIT  $\times$  \it BIT  $\times$  \it
BIT $\times$  \it BIT  $\times$  \it BIT\rm) $\fun$  \rm (\rm \{\it f0\rm \}  $\times$  \it BYTE\rm ).  This function is defined as:\\

@@

\subsection{Modeling the actions and instructions}
%TODO: Fiz uma breve alteração
Each instruction is represented by a B operation in Z80 module. The main
module (\textit{Z80}) has two categories of operations: a category represents the 
instructions of microcontrollers and a category represents the external actions, e.g.
interruptions actions. All 158 instructions and interruptions were modeled, but only
two important examples are presented here.

A simple example of instruction is $\textit{LD\_(nn)\_A}$\footnote{The B tools does not allow o use parentheses in identifiers, and the characters \{``('',``)''\} are
replaced respectively by \{``9'',``0''\} in the actual specification.} shown below. The
pre-defined functions are necessary many times to model the instructions, these
functions facilitate the construction of instruction set model. By default, all
parameters in operations are either predefined elements in the model or
integers values in decimal representation. This instruction use the
$\textit{updateAddressMem}$ operation from \textit{Memory} module, it receives
an address memory and its new memory value. $\textit{LD\_(nn)\_A}$ also increments the program
counter ($\textit{pc}$) and updates the refresh register ($\textit{r\_}$). The other instructions have a similar structure.

@@

% fundamenta melhor o trabalho já desenvolvido. Citar o artigo SBMF e a dissertação. 
% Os meus trabalhos já desenvolvidos/ resultados atuais.
The following works are directly related to the objective this paper.
The approach to develop verified software down to the assembly level
using B was described in~\cite{DantasSemish2008}. A first case study
for this approach was reported in~\cite{Dantas_SBMF08}, presenting
more details as well as a small software that developed up to assembly
level in three different platforms. A general view of a previous used
in verification process were presented in~\cite{Valerio_SBMF09}. The
model presented in the current paper added support for animation in
ProB and the representation of external actions (interruptions).
%Link com o parágrafo anterior 
This set of papers present an important step to software verification
up to assembly language, showing the current difficulties and
suggesting improvements in tools supporting the B method.

>>>>

\title{Experience in Modeling a Microcontroller Instruction Set Using B}
%
\hyphenation{se-pa-ra-te}
\hyphenation{in-struc-tion}
\hyphenation{re-gis-ter}
\hyphenation{re-gis-ters}
%\hyphenation{ge-ne-ra-ted}

@@

  This paper presents a formal model, developed in B, of the Z80
  instruction set. The formal model includes different libraries that
  specify entities commonly used in microcontrolers and
  microprocessors. Not only this model can be used for documentation
  purposes, but also to simulate and test the assembly code. It is also
  a fundamental component to extend the reach of the B method from the
  specification of the functional requirements of a system to an assembly-level
  implementation thereof.\newline %\keywords{Embedded Software, Simulation, B method and Verification}
% Using this model-driven approach and the instruction set model, a small case
% study to analyse a petroleum production test system in oil field was developed.
 { \textbf{Key words:} B method, Simulation, Verification and Z80}

\end{abstract}

@@

The new Z80 model supports animation in ProB~\cite{proB}.  Essentially, the
new model\footnote{The interested reader in more details is invited to visit
our repository at: http://code.google.com/p/b2asm/.} no longer contains
definitions derived from infinite sets. Using B constructs for modularity,
auxiliary libraries of basic modules were developed as part of the construction
of the microcontroller model. Such a library has many definitions of 
concepts common to microcontrollers. Besides the new Z80 model, a petroleum
production test system was developed using the Z80 assembly language to
analyse the code simulation and the verification process.

@@

Similarly, several other general functions and lemmas were created for all other data types.
Some times, we need to create arithmetic and logic functions that are more specific for a
determined microprocessor, like Z80 that is presented in the section \ref{sec:z80}.

The definitions, functions and properties from hardware library and types library are difficult to verify, otherwise these libraries can be reused in several others microprocessors without to verify again. Nearly 50\% of all these proof obligations shown in the Table~\ref{tab:types} are verified automatically or with few user pass\footnote{The user pass allows to filter proof tactics that aid the prover to find the proof, this is a feature provided by Atelier B.}. But many others proof obligations are difficult to verify because they involve functions that contain complex arithmetic expressions to manipulate with the interactive prover of AtelierB. However, all these 531 proof obligations were verified completely.

@@

\begin{figure}[he]
\centering
\includegraphics[width=2.8in]{images/Diagram_Types_and_Rules.pdf}
\caption{Conversion functions}
\label{DiagramTypesRules}
\end{figure}

@@

There is a specific function to refresh the flag register, it is named $\textit{update\_reg}$ $\textit{\_flag}$. It is typed and defined as follow:\it update\_flag\_reg \rm $\in$ \rm (\it BIT  $\times$  \it BIT  $\times$  \it BIT  $\times$  \it
BIT $\times$  \it BIT  $\times$  \it BIT\rm) $\fun$  \rm (\rm \{\it f0\rm \}  $\times$  \it BYTE\rm ).  This function is defined as:\\

@@

\subsection{Modeling the actions and instructions}
%TODO: Fiz uma breve alteração
Each instruction is represented by a B operation in Z80 module. The main
module (\textit{Z80}) has two categories of operations: a category represents the 
instructions of microcontrollers and a category represents the external actions, e.g.
interruptions actions. All 158 instructions and interruptions were modeled, but only
two important operations are presented here.

A simple example of instruction is $\textit{LD\_(nn)\_A}$\footnote{The B tools does not allow o use parentheses in identifiers, and the characters \{``('',``)''\} are
replaced respectively by \{``9'',``0''\} in the actual specification.} shown in this section. The
pre-defined functions are necessary many times to model the instructions, these
functions facilitate the construction of instruction set model. By default, all
parameters in operations are either predefined elements in the model or
integers values in decimal representation. This instruction use the
$\textit{updateAddressMem}$ operation from \textit{Memory} module, it receives
an address memory and its new memory value. $\textit{LD\_(nn)\_A}$ also increments the program
counter ($\textit{pc}$) and updates the refresh register ($\textit{r\_}$). The other instructions have a similar structure.

@@

% fundamenta melhor o trabalho já desenvolvido. Citar o artigo SBMF e a dissertação. 
% Os meus trabalhos já desenvolvidos/ resultados atuais.
The following works are directly related to the objective this paper.
The approach to develop verified software down to the assembly level
using B was described in~\cite{DantasSemish2008}. A first case study
for this approach was reported in~\cite{Dantas_SBMF08}, presenting
more details as well as a small software that developed up to assembly
level in three different platforms. A short general view of a previous work
were presented in~\cite{Valerio_SBMF09}. The model presented in the 
current paper added support for animation in ProB and the representation
of external actions (interruptions).
%Link com o parágrafo anterior 
This set of papers present an alternative step to software verification
up to assembly language, showing the current difficulties and
suggesting improvements in tools supporting the B method.
